Explicit B-spline regularization in diffeomorphic image registration

### 2013-10-28 paper.tex

\begin{document}
\onecolumn
\firstpage{1}

@@

\begin{abstract}
Important methodological developments in the evolution of image registration algorithms include those
in which the correspondence relationship is characterized by diffeomorphisms.
The  popularity of these approaches is due largely to their
topological properties and success in providing biologically plausible
solutions to deformation and morphological
estimation problems. Popular variants of diffeomorphic algorithms include those characterized by
time-varying and constant velocity fields, and symmetrical considerations.
Prior information in the form of regularization is used to enforce transform plausibility taking the form of physics-based constraints or through some approximation thereof, e.g. Gaussian smoothing of the vector fields (a la Thirion's Demons \citep{thirion1998}).  In the context of the original Demons' framework, the {\it directly manipulated free-form deformation} approach \citep{tustison2009}, can be viewed as a smoothing alternative in which explicit regularization is achieved through fast B-spline approximation.
This characterization can be used to provide B-spline ``flavored'' diffeomorphic image registration solutions with several advantages.  Implementation is open source and available through the Insight Toolkit and our Advanced Normalization Tools (ANTs) repository.  A thorough comparative evaluation with the well-known SyN algorithm \citep{avants2008}, implemented within the same framework, and its B-spline analog is performed using open labeled brain data and open source evaluation tools.  
\tiny
\keyFont{ \section{Keywords:}
Advanced Normalization Tools, diffeomorphisms, directly manipulated free-form deformation, Insight Toolkit, spatial normalization }
\end{abstract}

@@

Establishment of anatomical and functional correspondence
is a crucial step towards gaining insight into certain biological
sciences.  Neuroscience research efforts, such as characterizing
brain morphology, require accurate and robust methods for
producing such mappings.  The
extensive literature detailing methodology is evidence of the rich history of
algorithmic development which continues contemporaneously.
We highlight several key historical contributions which are particularly relevant to the work 
subsequently presented.

@@

It is interesting to note that within this context, traditional FFD algorithms
can be viewed as a type of fluid-like Demons approach
where, rather than projecting the update field to the space
of regularized fields using Gaussian convolution, gradient fields
are projected to a smooth space characterized by the B-spline
basis functions.  This analogy was hinted at in our earlier work \citep{tustison2009} where we showed that fitting the update field to a B-spline object using a fast approximation routine \citep{tustison2006} is equivalent to a preconditioning of the standard gradient used in gradient descent-based FFD optimization.
This preconditioning is used to mitigate the hemstitching effect induced by the ill-conditioned nature of the traditional gradient-based FFD formulation.  We denote
this new FFD variant as directly manipulated free-form deformation (DMFFD) and, 
as part of the ITKv4 refactoring efforts, has been implemented for use with the new
registration framework%
\footnote{
http://www.itk.org/Doxygen/html/classitk\_1\_1BSplineSmoothingOnUpdateDisplacementFieldTransform.html
}
which permits both B-spline smoothing on the update (``viscous'') and total (``elastic'')
displacement fields at each iteration (cf
analogous Gaussian, i.e. Demons, implementation%
\footnote{
http://www.itk.org/Doxygen/html/classitk\_1\_1GaussianSmoothingOnUpdateDisplacementFieldTransform.html
}).

Continuing from the work of \cite{christensen1996} and subsequent exploration into the mathematical formalisms of diffeomorphisms \citep[e.g.][]{dupuis1998},
the well-known Large Deformation Diffeomorphic Metric Mapping (LDDMM) algorithm
was proposed in \cite{beg2005}.  In contrast to the mapping produced
by \cite{christensen1996}, LDDMM yields the geodesic solution in the space of diffeomorphisms between two images. Since its introduction, LDDMM has
inspired much innovation in the image registration community.  Applying
the log-Euclidean framework of \cite{arsigny2006}, DARTEL (Diffeomorphic Anatomic Registration using Exponential Lie algebra) uses a constant velocity field parameterization to provide a fast, diffeomorphic alternative \citep{ashburner2007}. Additionally, symmetrical considerations in the velocity field parameterization are discussed in \cite{avants2008} in the context of a cross correlation similarity metric.
By explicitly symmetrizing the LDDMM formulation, this Symmetric Normalization (SyN) minimizes the bias of the resulting transformation when selecting the ``fixed'' and ``moving'' images.  A greedy version of this algorithm has proven successful in neuroimaging \citep{klein2009} and pulmonary \citep{murphy2011} applications.

@@

Therefore, in addition to the full-scale
symmetric SyN offering described in \cite{avants2008}, the authors therein
provided a ``greedy'' alternative which has demonstrated superior performance
in various applications \citep{avants2011,klein2009,murphy2011}%
\footnote{
SyN was also the standard registration for the MICCAI 2013
Workshop on Segmentation:  Algorithms Theory, and Applications (SATA)
and corresponding challenge (https://masi.vuse.vanderbilt.edu/workshop2013).
}
while simultaneously being capable of running with limited computational
resources.  This is due to the restriction of the discrete, time-parameterized
velocity field samples to their respective
endpoints, i.e. the $v_i(\mathbf{x},t)$ are sampled at $t \in \{0,0.5,1\}$
implying simultaneous storage of only four transform vector fields
$\phi_1(\mathbf{x})$,
$\phi_1^{-1}(\mathbf{x})$, $\phi_2(\mathbf{x})$, and $\phi_2^{-1}(\mathbf{x})$
(cf Figure \ref{fig:syn}).  Since the greedy SyN framework is the focus of the evaluation,
the algorithmic steps are briefly sketched
in Algorithm \ref{alg:syn}.

@@

>>>>

%\selectlanguage{eng}
\begin{document}
\onecolumn
\firstpage{1}

@@

\begin{abstract}
Diffeomorphic maps are central to image registration due largely to their
topological properties and success in providing biologically plausible
solutions to deformation and morphological
estimation problems. Popular diffeomorphic algorithms include those characterized by
time-varying and constant velocity fields, and symmetrical considerations.
Prior information in the form of regularization is used to enforce transform plausibility 
taking the form of physics-based constraints or through some approximation thereof, e.g. 
Gaussian smoothing of the vector fields (a la Thirion's Demons \citep{thirion1998}).  In 
the context of the original Demons' framework, the so-called {\it directly manipulated free-form deformation} \citep{tustison2009}
can be viewed as a smoothing alternative  in which explicit regularization is achieved through 
fast B-spline approximation.
This characterization can be used to provide B-spline ``flavored'' diffeomorphic image 
registration solutions with several advantages.  Implementation 
is open source and available through the Insight Toolkit and our Advanced Normalization Tools
(ANTs) repository.  A thorough comparative evaluation with the well-known SyN algorithm 
\citep{avants2008}, implemented within the same framework, and its B-spline analog is 
performed using open labeled brain data and open source evaluation tools.
\tiny
\keyFont{ \section{Keywords:}
Advanced Normalization Tools, diffeomorphisms, directly manipulated free-form deformation, Insight Toolkit, spatial normalization }
\end{abstract}

@@

Establishment of anatomical and functional correspondence is a crucial
step towards gaining insight into biological processes.
Neuroscience research efforts, such as characterizing brain
morphology, require accurate and robust methods for producing such
mappings.  The extensive literature detailing methodology is evidence
of the rich history of algorithmic development which continues
contemporaneously.  We highlight several key historical contributions
which are particularly relevant to the work presented.

@@

It is interesting to note that within this context, traditional FFD
algorithms can be viewed as a type of fluid-like Demons approach
where, rather than projecting the update field to the space of
regularized fields using Gaussian convolution, gradient fields are
projected to a smooth space characterized by the B-spline basis
functions.  This analogy was hinted at in our earlier work
\citep{tustison2009} where we showed that fitting the update field to
a B-spline object using a fast approximation routine
\citep{tustison2006} is equivalent to a preconditioning of the
standard gradient used in gradient descent-based FFD optimization.
This preconditioning is used to mitigate the hemstitching effect
induced by the ill-conditioned nature of the traditional
gradient-based FFD formulation.  We denoted this new FFD variant as
directly manipulated free-form deformation (DMFFD) and, as part of the
ITKv4 refactoring efforts, has been implemented for use with the new
registration framework%
\footnote{
http://www.itk.org/Doxygen/html/classitk\_1\_1BSplineSmoothingOnUpdateDisplacementFieldTransform.html
}
which permits both B-spline smoothing on the update (``viscous'') and total (``elastic'')
displacement fields at each iteration (cf
analogous Gaussian, i.e. Demons, implementation%
\footnote{
http://www.itk.org/Doxygen/html/classitk\_1\_1GaussianSmoothingOnUpdateDisplacementFieldTransform.html
}).

Continuing from the work of \cite{christensen1996} and subsequent
exploration into the mathematical formalisms of diffeomorphisms
\citep[e.g.][]{dupuis1998}, the well-known Large Deformation
Diffeomorphic Metric Mapping (LDDMM) algorithm was proposed in
\cite{beg2005}.  In contrast to the mapping produced by
\cite{christensen1996}, LDDMM yields the geodesic solution in the
space of diffeomorphisms between two images. Since its introduction,
LDDMM has inspired much innovation in the image registration
literature.  Applying the log-Euclidean framework of
\cite{arsigny2006}, DARTEL (Diffeomorphic Anatomic Registration using
Exponential Lie algebra) uses a constant velocity field
parameterization to provide a fast, diffeomorphic alternative
\citep{ashburner2007}. Additionally, symmetrical considerations in the
velocity field parameterization are discussed in \cite{avants2008} in
the context of a cross correlation similarity metric.  By explicitly
symmetrizing the LDDMM formulation, this Symmetric Normalization (SyN)
minimizes the bias of the resulting transformation when selecting the
``fixed'' and ``moving'' images.  A greedy version of this algorithm
has proven successful in neuroimaging \citep{klein2009} and pulmonary
\citep{murphy2011} applications as well as in multi-atlas label fusion \cite{wang2012}.

@@

Therefore, in addition to the full-scale
SyN offering described in \cite{avants2008}, the authors therein
provided a ``greedy'' alternative which has demonstrated superior performance
in differents applications \citep{avants2011,klein2009,murphy2011}%
\footnote{
SyN was also the standard registration for the MICCAI 2013
Workshop on Segmentation:  Algorithms Theory, and Applications (SATA)
and corresponding challenge (https://masi.vuse.vanderbilt.edu/workshop2013).
}
while simultaneously being capable of running with limited computational
resources.  This is due to the restriction of the discrete, time-parameterized
velocity field samples to their respective
endpoints, i.e. the $v_i(\mathbf{x},t)$ are sampled at $t \in \{0,0.5,1\}$
implying simultaneous storage of only four transform vector fields
$\phi_1(\mathbf{x})$,
$\phi_1^{-1}(\mathbf{x})$, $\phi_2(\mathbf{x})$, and $\phi_2^{-1}(\mathbf{x})$
(cf Figure \ref{fig:syn}).  Furthermore, the forward and inverse
mappings are guaranteed to be consistent within the discrete domain i.e. $\|
\phi_i^{-1}(\phi_i) - \mathbf{Id} \|^2 < \epsilon$.   
Since the greedy SyN framework is the focus of the evaluation,
the algorithmic steps are briefly sketched
in Algorithm \ref{alg:syn}.

@@

A significant amount of research has been devoted to image registration
algorithmic development.  Given their many salient characteristics particularly with
respect to large deformation estimation constrained by topological 
continuity, diffeomorphic registration
approaches have been a particular focus in the neuroimaging community.
However, many groups continue to find success with non-diffeomorphic
FFD methods
\citep[e.g.][]{rueckert1999,klein2010}.  Using our DMFFD framework, 
B-spline regularization is easily adapted into the diffeomorphic registration 
framework and performs comparably to analogous algorithms which we demonstrated 
in this work for the case of the widely-used SyN.
